For Induction-Induction, Induction is Enough

### 2019-11-11 paper/paper.tex
small fixes, lot more to come

The current work builds heavily on the work of Kaposi et al.\ on quotient
inductive-inductive types\cite{Kaposi:2019:CQI:3302515.3290315}; we reuse both
QIIT syntax and semantics by restricting to IITs, and we reuse the term model
construction of QIITs as well.

Forsberg\cite{forsberg-phd} specifies inductive-inductive types and develops a
semantics based on dialgebras. He also provides a reduction of IITs to indexed
inductive types, but only constructs a weak version of elimination.

Hugunin\cite{jasper} constructs several IITs in cubical Agda from inductive
types. In this setting, the lack of UIP makes constructions significantly more
involved, and essentially involves coinductive-coinductive well-formedness
predicates defined as homotopy limits. Hugunin does not consider a generic
syntax of IITs and only works on specific examples (although the examples vary
greatly). He also does not construct full recursive-recursive eliminators.

Streicher\cite{streichersemantics} presents an interpretation of the well-formed
presyntax of a type theory into a categorical model, which is an important
ingredient in constructing an initial model, although he does not present
details on the construction of the term model or its initiality proof.

Voevodsky was interested in constructing initial models of type theories
from presyntaxes. There is a communal project \cite{

@@

Agda formalisation of Sections \ref{sec:theory_of_signatures} and
\ref{sec:ambroise} can be found at
\url{https://github.com/amblafont/UniversalII}. In addition to the features in
the previous section, this formalisation also uses equality reflection for some
equalities, through rewriting pragmas\cite{cockxsprinkles}. This is purely for
convenience, and in principle we could eliminate the instances of equality
reflection via UIP and function extensionality.

@@

The theory of IIT signatures is given as a category with families
(CwF)\cite{Dybjer96internaltype} with certain additional type formers. The CwF
part constitutes the substitution calculus, while the type formers allow
declaring new sorts and operators in a signature. Below, we list all components
for the theory of signatures; this can be viewed as a notion of model for the
theory, which is an iterated $\Sigma$ type. In this paper, we construct the
\emph{initial} such model, i.e.\ a model equipped with a unique recursion principle.

@@

\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \Con && : \Set \\
  & \Ty  && : \Con\ra\Set \\
  & \Sub  && : \Con\ra\Con\ra\Set \\
  & \Tm  && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Set \\
  & \id && : \Sub\,\Gamma\,\Gamma \\
  & \blank\circ\blank && : \Sub\,\Theta\,\Delta\ra\Sub\,\Gamma\,\Theta\ra\Sub\,\Gamma\,\Delta \\
  & \ass && : (\sigma \circ \delta) \circ \nu = \sigma \circ (\delta \circ \nu) \\
  & \idl && : \id\circ\sigma = \sigma \\
  & \idr && : \sigma\circ\id = \sigma \\
  & \blank[\blank] && : \Ty\,\Delta\ra\Sub\,\Gamma\,\Delta\ra\Ty\,\Gamma \\
  & \blank[\blank] && : \Tm\,\Delta\,A\ra(\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma]) \\
  & [\id] && : A[\id] = A \\
  & [\circ] && : A[\sigma \circ \delta] = A[\sigma][\delta] \\
  & [\id] && : t[\id] = t \\
  & [\circ] && : t[\sigma \circ \delta] = t[\sigma][\delta] \\
  & \boldsymbol{\cdot} && : \Con \\
  & \epsilon && : \Sub\,\Gamma\,\boldsymbol{\cdot} \\
  & {\boldsymbol{\cdot}\eta} && : \{\sigma : \Sub\,\Gamma\,\boldsymbol{\cdot}\} \ra \sigma = \epsilon \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \blank,\blank && : (\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma])\ra\Sub\,\Gamma\,(\Delta\ext A) \\
  & \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
  & \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
  & {\ext\beta_1} && : \p\circ(\sigma, t) = \sigma \\
  & {\ext\beta_2} && : \q[\sigma, t] = t \\
  & {\ext\eta} && : (\p, \q) = \id \\
  & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\[0.5em]
  & \rlap{(2) Universe} \\[0.5em]
  & \U && : \Ty\,\Gamma \\
  & \El && : \Tm\,\Gamma\,\U\ra\Ty\,\Gamma \\
  & {\U[]} && : \U[\sigma] = \U \\
  & {\El[]} && : (\El\,a)[\sigma] = \El\,(a[\sigma]) \\[0.5em]
  & \rlap{(3) Inductive parameters} \\[0.5em]
  & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\ext\El\,a)\ra\Ty\,\Gamma \\
  & \app && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra\Tm\,(\Gamma\ext \El\,a)\,B \\
  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma^\uparrow]) \\
  & {\app[]} && : (\app\,t)[\sigma^\uparrow] = \app\,(t[\sigma]) \\[0.5em]
  & \rlap{(4) Metatheoretic parameters} \\[0.5em]
  & \Pim && : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma \\
  & \blank\appm \blank && : \Tm\,\Gamma\,(\Pim\,T\,B)\ra(\alpha:T) \ra\Tm\,\Gamma\,(B\,\alpha) \\
  & {\Pim[]} && : (\Pim\,T\,B)[\sigma] = \Pim\,T\,(\lambda \alpha.(B\,\alpha)[\sigma]) \\
  & {\appm[]} && : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha \\[0.5em]
\end{alignat*}
\subsubsection{Notation and abbreviations}

We define $\sigma^\uparrow$ as $(\sigma\circ\p,\q)$. In the $\Pi$ function type
for inductive parameters, we have $\app$ as a categorical version of function
application. We recover the usual type-theoretic application as follows:
\begin{alignat*}{5}
  & \blank\oldapp\blank : \Tm\,\Gamma\,(\Pi\,a\,B)\ra (u : \Tm\,\Gamma\,(\El\,a))
  \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
  & t \oldapp u := \app\,t\,[\id\,u]
\end{alignat*}
We also abbreviate non-dependent $\Pi$ functions by writing $\Pi\,a\,(B[p])$ as $a\ra B$.

We can also recover de Bruijn indices by having $n := q[p^n]$, where $p^n$ is
$p$ composed with itself $n$ times. However, for informal examples we instead
use a nameful notation, where we introduce names in context extension and in the
$\Pi$ function domain, for example as $\boldsymbol{\cdot}\,\ext\,A : \U\,\ext\,a :
\El\,A$ in a signature for pointed sets.

\subsubsection{Example signatures}

Natural numbers:
\[
\boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \ra \El\,N
\]
Lists with elements of a given $A : \Set$ type. Here we use the $\Pim$ function
to include an external type into the signature.
\[
\boldsymbol{\cdot}\,\ext\,L:\U\,\ext\,nil : \El\,L\,\ext\,
      cons : \Pim\,A\,(\lambda \_.\, L\ra L)
\]
The inductive-inductive example from Section \ref{sec:intro}:
\begin{alignat*}{3}
  & \boldsymbol{\cdot} \,\ext\,Con : \U\,\ext\,Ty : Con\ra\U\,\ext\,empty : \El\,Con\,\ext\,ext : \Pi\,(\Gamma : Con)\,(Ty\oldapp\Gamma \ra Con) \\
  & \hspace{0.5em}\,\ext\, U : \Pi\,(\Gamma:Con)\,(\El\,(Ty\oldapp\Gamma))
    \,\ext\,El : \Pi\,(\Gamma : Con)(\El\,(Ty\oldapp(ext\oldapp\Gamma\oldapp (U\oldapp\Gamma))))
\end{alignat*}

@@

However, we shall mention that the Agda formalisation for the current paper
cannot be directly plugged into the Agda code for
\cite{Kaposi:2019:CQI:3302515.3290315} (which includes most of the term model
construction). One reason is that the QIIT formalisation uses strict computation
rules (given by Agda rewrite rules) for induction over signatures, while here we show propositional
computation. This mismatch can be in principle remedied by noting that both
formalisations use UIP and function extensionality, hence we can switch between
strict and propositional equalities, via the known translations between
extensional and intensional type
theories\cite{hofmann95extensional,winterhalter2019eliminating}.

>>>>

The current work builds heavily on the work of Kaposi et al.\ on quotient
inductive-inductive types \cite{Kaposi:2019:CQI:3302515.3290315}; we reuse both
QIIT syntax and semantics by restricting to IITs, and we reuse the term model
construction of QIITs as well.

Forsberg \cite{forsberg-phd} specifies inductive-inductive types and develops a
semantics based on dialgebras. He also provides a reduction of IITs to indexed
inductive types, but only constructs a weak version of elimination.

Hugunin \cite{jasper} constructs several IITs in cubical Agda from inductive
types. In this setting, the lack of UIP makes constructions significantly more
involved, and essentially involves coinductive-coinductive well-formedness
predicates defined as homotopy limits. Hugunin does not consider a generic
syntax of IITs and only works on specific examples (although the examples vary
greatly). He also does not construct full recursive-recursive eliminators.

Streicher \cite{streichersemantics} presents an interpretation of the well-formed
presyntax of a type theory into a categorical model, which is an important
ingredient in constructing an initial model, although he does not present
details on the construction of the term model or its initiality proof.

Voevodsky was interested in constructing initial models of type theories
from presyntaxes. There is a communal project \cite{}

@@

Agda formalisation of Sections \ref{sec:theory_of_signatures} and
\ref{sec:ambroise} can be found at
\url{https://github.com/amblafont/UniversalII}. In addition to the features in
the previous section, this formalisation also uses equality reflection for some
equalities, through rewriting pragmas \cite{cockxsprinkles}. This is purely for
convenience, and in principle we could eliminate the instances of equality
reflection via UIP and function extensionality.

@@

The theory of IIT signatures is given as a category with families
(CwF) \cite{Dybjer96internaltype} with certain additional type formers. The CwF
part constitutes the substitution calculus, while the type formers allow
declaring new sorts and operators in a signature. Below, we list all components
for the theory of signatures; this can be viewed as a notion of model for the
theory, which is an iterated $\Sigma$ type. In this paper, we construct the
\emph{initial} such model, i.e.\ a model equipped with a unique recursion principle.

@@

\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \Con && : \Set \\
  & \Ty  && : \Con\ra\Set \\
  & \Sub  && : \Con\ra\Con\ra\Set \\
  & \Tm  && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Set \\
  & \id && : \Sub\,\Gamma\,\Gamma \\
  & \blank\circ\blank && : \Sub\,\Theta\,\Delta\ra\Sub\,\Gamma\,\Theta\ra\Sub\,\Gamma\,\Delta \\
  & \ass && : (\sigma \circ \delta) \circ \nu = \sigma \circ (\delta \circ \nu) \\
  & \idl && : \id\circ\sigma = \sigma \\
  & \idr && : \sigma\circ\id = \sigma \\
  & \blank[\blank] && : \Ty\,\Delta\ra\Sub\,\Gamma\,\Delta\ra\Ty\,\Gamma \\
  & \blank[\blank] && : \Tm\,\Delta\,A\ra(\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma]) \\
  & [\id] && : A[\id] = A \\
  & [\circ] && : A[\sigma \circ \delta] = A[\sigma][\delta] \\
  & [\id] && : t[\id] = t \\
  & [\circ] && : t[\sigma \circ \delta] = t[\sigma][\delta] \\
  & \boldsymbol{\cdot} && : \Con \\
  & \epsilon && : \Sub\,\Gamma\,\boldsymbol{\cdot} \\
  & {\boldsymbol{\cdot}\eta} && : \{\sigma : \Sub\,\Gamma\,\boldsymbol{\cdot}\} \ra \sigma = \epsilon \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \blank,\blank && : (\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma])\ra\Sub\,\Gamma\,(\Delta\ext A) \\
  & \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
  & \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
  & {\ext\beta_1} && : \p\circ(\sigma, t) = \sigma \\
  & {\ext\beta_2} && : \q[\sigma, t] = t \\
  & {\ext\eta} && : (\p, \q) = \id \\
  & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\[0.5em]
  & \rlap{(2) Universe} \\[0.5em]
  & \U && : \Ty\,\Gamma \\
  & \El && : \Tm\,\Gamma\,\U\ra\Ty\,\Gamma \\
  & {\U[]} && : \U[\sigma] = \U \\
  & {\El[]} && : (\El\,a)[\sigma] = \El\,(a[\sigma]) \\[0.5em]
  & \rlap{(3) Inductive parameters} \\[0.5em]
  & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\ext\El\,a)\ra\Ty\,\Gamma \\
  & \app && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra\Tm\,(\Gamma\ext \El\,a)\,B \\
  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma^\uparrow]) \\
  & {\app[]} && : (\app\,t)[\sigma^\uparrow] = \app\,(t[\sigma]) \\[0.5em]
  & \rlap{(4) Metatheoretic parameters} \\[0.5em]
  & \Pim && : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma \\
  & \blank\appm \blank && : \Tm\,\Gamma\,(\Pim\,T\,B)\ra(\alpha:T) \ra\Tm\,\Gamma\,(B\,\alpha) \\
  & {\Pim[]} && : (\Pim\,T\,B)[\sigma] = \Pim\,T\,(\lambda \alpha.(B\,\alpha)[\sigma]) \\
  & {\appm[]} && : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha \\[0.5em]
\end{alignat*}
\subsection{Notation and abbreviations}

We define $\sigma^\uparrow$ as $(\sigma\circ\p,\q)$. In the $\Pi$ function type
for inductive parameters, we have $\app$ as a categorical version of function
application. We recover the usual type-theoretic application as follows:
\begin{alignat*}{5}
  & \blank\oldapp\blank : \Tm\,\Gamma\,(\Pi\,a\,B)\ra (u : \Tm\,\Gamma\,(\El\,a))
  \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
  & t \oldapp u := (\app\,t)[\id\,u]
\end{alignat*}
We also abbreviate non-dependent $\Pi$ functions by writing $\Pi\,a\,(B[\p])$ as $a\Ra B$.

We can also recover de Bruijn indices by having $n := \q[\p^n]$, where $\p^n$ is
$\p$ composed with itself $n$ times. However, for informal examples we instead
use a nameful notation, where we introduce names in context extension and in the
$\Pi$ function domain, for example as $\boldsymbol{\cdot}\,\ext\,A : \U\,\ext\,a :
\El\,A$ in a signature for pointed sets.

\subsection{Example signatures}

Natural numbers:
\[
\boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N
\]
Lists with elements of a given $A : \Set$ type. Here we use the $\Pim$ function
to include an external type into the signature.
\[
\boldsymbol{\cdot}\,\ext\,L:\U\,\ext\,nil : \El\,L\,\ext\,
      cons : \Pim\,A\,(\lambda \_.\, L\Ra L)
\]
The inductive-inductive example from Section \ref{sec:intro}:
\begin{alignat*}{3}
  & \boldsymbol{\cdot} \,\ext\,Con : \U\,\ext\,Ty : Con\Ra\U\,\ext\,empty : \El\,Con\,\ext\,ext : \Pi\,(\Gamma : Con)\,(Ty\oldapp\Gamma \Ra Con) \\
  & \hspace{0.5em}\,\ext\, U : \Pi\,(\Gamma:Con)\,(\El\,(Ty\oldapp\Gamma))
    \,\ext\,El : \Pi\,(\Gamma : Con)(\El\,(Ty\oldapp(ext\oldapp\Gamma\oldapp (U\oldapp\Gamma))))
\end{alignat*}

@@

However, we shall mention that the Agda formalisation for the current paper
cannot be directly plugged into the Agda code for
\cite{Kaposi:2019:CQI:3302515.3290315} (which includes most of the term model
construction). One reason is that the QIIT formalisation uses strict computation
rules (given by Agda rewrite rules) for induction over signatures, while here we show propositional
computation. This mismatch can be in principle remedied by noting that both
formalisations use UIP and function extensionality, hence we can switch between
strict and propositional equalities, via the known translations between
extensional and intensional type
theories \cite{hofmann95extensional,winterhalter2019eliminating}.

### 2019-11-11 paper/universaliit.tex
small fixes, lot more to come

Note that $(\Piu \,A\,B)[\sigma]$ should be defined as $\Piu
\,A[\sigma]\,B[\varu\, 0 \,\consu\, \wkO\,\sigma]$, and thus we need to define
$\wk0$, the weakening of substitutions.  The basic idea is to increment the de
Bruijn indices of all the variables.  Actually this is not so simple because of
the $\Piu$ type: indeed, we want to define $\wk(\Piu\,A\,B)$ as the $\Pi$ type
of the weakening of $A$ and $B$, but here, $B$ must be weakened with respect to
the second last variable of the context, rather than the last one.  For this
reason, we need to generalize the weakening as occuring anywhere in the context.
\begin{alignat*}{5}
  & \wk_n && :  \Tyu\ra\Tyu \\
  & \wk_n && :  \Tmu\ra\Tmu \\
  % & \wk && : \N \ra \N\ra\N \\
  & \wkO && : \Subu\ra \Subu
  \end{alignat*}
  The natural number $n$ specifies at which position of the context the
  weakening occurs.
  Here, $\wkO$ weakens with respect to the last variable.

Later, in Section~\ref{ss:weaken-typing}, we show that weakening preserves
typing. Stating a typing rule for this operation requires to have a way to
express the weakening occuring at the middle of a context. We consider pairs of
untyped contexts, which should be thought of as a splitting of a context at some
position. The full context is recovered by merging the two components:
% (which appears in the typing rule of a weakening occuring in the middle of a context),
% and the length $\length{\Gamma}$ of a
% context $\Gamma$:
\begin{alignat*}{5}
  & \blank \merge \blank && :  \Conu\ra\Conu\ra\Conu \\
  & \Gamma \merge \cdot && :=  \Gamma \\
  & \Gamma \merge (\Delta \extu A) && :=  (\Gamma \merge \Delta)\extu A
\end{alignat*}
We think of the second context as a telescope over the first one. We also define
weakening for telescopes, which will be used to give typing rules for telescopes
in Section \ref{sec:typing_judgments}:
\begin{alignat*}{5}
  & \wkO && : \Conu\ra \Conu
  \\
  & \wkO\,\emptyu && := \emptyu
  \\
  & \wkO\,(\Delta\,\extu A) && := \wkO\,\Delta\,\extu \wk_{\length{\Delta}}\,A
  \end{alignat*}
  \subsubsection{Untyped substitution}
  We define single substitution by recursion on presyntax:
  % & \rlap{(1) Substitution calculus} \\[0.5em]
\begin{alignat*}{5}
  & \blank[\blank := \blank] && : \Tyu\ra\N\ra\Tmu\ra\Tyu \\
  & \blank[\blank := \blank] && : \Tmu\ra\N\ra\Tmu\ra\Tmu
  % \\
  % & \blank[\blank := \blank] && : \N\ra\N\ra\Tmu\ra\Tmu
  \end{alignat*}
  This is enough to define the typing judgments: indeed, the typing rule for application
  involves only a unary substitution.

  However, to construct the initial model of the QIIT, we need to define
  the full substitution calculus:
\begin{alignat*}{5}
  & \blank[\blank] && : \Tyu\ra\Subu\ra\Tyu \\
  & \blank[\blank] && : \Tmu\ra\Subu\ra\Tmu \\
  & \blank \circ \blank && : \Subu\ra\Subu\ra \Subu
  \end{alignat*}
These can be defined either by iterating unary substitutions, or by recursion
on untyped syntax: the two ways yield provably equal definitions. In the
following, we assume that it is defined by recursion. We also make use of the
following definition:
  \begin{alignat*}{5}
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \keep && : \Subu \ra \Subu \\
   &&& := \lambda \sigma. \varu\, 0 \,\consu\, \wkO\,\sigma
  \end{alignat*}
  The idea is that if $\sigma$ is a substitution between contexts $\Gamma$ and
  $\Delta$, then $\keep\,\sigma$ is a substitution between contexts $\Gamma, A[\sigma]$
  and $\Delta,A$ for any type $A$. This occurs when defining
  $(\Piu\,A\,B)[\sigma]$ as $\Piu (A[\sigma])(B[\keep \,\sigma])$.

  We can define the identity substitution on a context $\Gamma$ as follows,
  where $\length{\Gamma}$ is the length of the context $\Gamma$, and
  $\keep^{\length{\Gamma}}$ is $\keep$ iterated $\length{\Gamma}$ times:
\begin{alignat*}{5}
  & \length{\Gamma} && : \Conu\ra\N
  \\
  & \idu && : \Conu \ra \Subu \\
  & && := \lambda \Gamma.\keep^{\length{\Gamma}} \epsilonu
  \end{alignat*}

@@

  \subsubsection{Typing judgments}
  \label{sec:typing_judgments}
  The typing judgments are defined as the following inductive datatype indexed over the
  untyped syntax:
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \blank \vdash && : \Conu\ra \Set \\
  & \blank \vdash \blank  && : \Conu\ra \Tyu\ra\Set \\
  & \blank \vdash \blank \in \blank  && : \Conu\ra \Tmu\ra \Tyu\ra \Set \\
  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
  & \blank \vdash \blank \Rightarrow \blank  && : \Conu\ra \Subu \ra \Conu\ra \Set \\
  % & \Subpre  && : \Set \\
  & \emptyw && : \emptyu \vdash \\
  & \epsilonw && : \Gamma\vdash \epsilonu \Rightarrow \emptyu \\
  & \blank\extw\blank && : ( \Gamma\vdash )  \ra  (\Gamma\vdash A)  \ra
  \Gamma\extu A \vdash \\
  & \consw && :
    (\Delta \vdash) \ra
    (\Gamma\vdash \sigma \Rightarrow\Delta)\ra
    ( \Delta \vdash A) \ra
    ( \Gamma \vdash t \in A[\sigma]) \ra
    \Gamma \vdash t \consu \sigma \Rightarrow \Delta\extu A
   \\
  & \varw  && : (\Gamma \vdash n \invar A) \ra \Gamma \vdash \varu n \in A \\
  & \vzw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra\Gamma \extu  A \vdash 0 \invar \wku \, A \\
  & \vsw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra(\Gamma \vdash n \invar A) \ra (\Gamma \vdash B) \ra \Gamma \extu  B
  \vdash S\,n \invar \wku \, A \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uw && : (\Gamma \vdash)\ra \Gamma \vdash \Uu \\
  & \Elw && : (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra\Gamma \vdash \Elu\,a \\
  & \rlap{(3) Inudctive parameters} \\[0.5em]
  & \Piw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
  \Elu\,a \vdash B)\ra\Gamma \vdash \Piu \, a \, B \\
  & \appw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
    \Elu\,a \vdash B)
    \\
    & &&
    \ra(\Gamma \vdash t \in \Piu \, a \, B )
    \ra
    (\Gamma \vdash u \in \Elu\,a)
    \ra
    \Gamma \vdash t \appu u \in  B [0 := u] \\
  & \rlap{(4) Inductive parameters} \\[0.5em]
  & \Pimw && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t) \ra
    \Gamma \vdash \Pimu \,T\,A
    \\
    & \appmw  && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t)
    \\
    & &&
    \ra(\Gamma \vdash t \in \Pimu \,T\,A)
    \ra (u : T) \ra \Gamma \vdash t\,\appmu\,u \in A\,u
\end{alignat*}
There is possibility of redundancy in the arguments of the constructors. Here,
we are ``{paranoid}'', so that we get more inductive hypotheses when performing
recursion.

>>>>

Note that $(\Piu \,A\,B)[\sigma]$ should be defined as $\Piu
\,(A[\sigma])\,(B[\varu\, 0 \,\consu\, \wkO\,\sigma])$, and thus we need to define
$\wkO$, the weakening of substitutions.  The basic idea is to increment the de
Bruijn indices of all the variables.  Actually this is not so simple because of
the $\Piu$ type: indeed, we want to define $\wkO\,(\Piu\,A\,B)$ as the $\Pi$ type
of the weakening of $A$ and $B$, but here, $B$ must be weakened with respect to
the second last variable of the context, rather than the last one.  For this
reason, we need to generalize the weakening as occuring anywhere in the context.
\begin{alignat*}{5}
  & \wk_n && :  \Tyu\ra\Tyu \\
  & \wk_n && :  \Tmu\ra\Tmu \\
  % & \wk && : \N \ra \N\ra\N \\
  & \wkO && : \Subu\ra \Subu
  \end{alignat*}
  The natural number $n$ specifies at which position of the context the
  weakening occurs.
  Here, $\wkO$ weakens with respect to the last variable.

Later, in Section~\ref{ss:weaken-typing}, we show that weakening preserves
typing. Stating a typing rule for this operation requires to have a way to
express the weakening occuring at the middle of a context. We consider pairs of
untyped contexts, which should be thought of as a splitting of a context at some
position. The full context is recovered by merging the two components:
% (which appears in the typing rule of a weakening occuring in the middle of a context),
% and the length $\length{\Gamma}$ of a
% context $\Gamma$:
\begin{alignat*}{5}
  & \blank \merge \blank && :  \Conu\ra\Conu\ra\Conu \\
  & \Gamma \merge \cdot && :=  \Gamma \\
  & \Gamma \merge (\Delta \extu A) && :=  (\Gamma \merge \Delta)\extu A
\end{alignat*}
We think of the second context as a telescope over the first one. We also define
weakening for telescopes, which will be used to give typing rules for telescopes
in Section \ref{sec:typing_judgments}:
\begin{alignat*}{5}
  & \wkO && : \Conu\ra \Conu
  \\
  & \wkO\,\emptyu && := \emptyu
  \\
  & \wkO\,(\Delta\,\extu A) && := \wkO\,\Delta\,\extu \wk_{\length{\Delta}}\,A
  \end{alignat*}
$\length{\Gamma}$ denotes the length of a the context $\Gamma$.

\subsubsection{Untyped substitution}
  We define single substitution by recursion on presyntax:
  % & \rlap{(1) Substitution calculus} \\[0.5em]
\begin{alignat*}{5}
  & \blank[\blank := \blank] && : \Tyu\ra\N\ra\Tmu\ra\Tyu \\
  & \blank[\blank := \blank] && : \Tmu\ra\N\ra\Tmu\ra\Tmu
  % \\
  % & \blank[\blank := \blank] && : \N\ra\N\ra\Tmu\ra\Tmu
  \end{alignat*}
  This is enough to define the typing judgments: indeed, the typing rule for application
  involves only a unary substitution.

  However, to construct the initial model of the QIIT, we need to define
  the full substitution calculus:
\begin{alignat*}{5}
  & \blank[\blank] && : \Tyu\ra\Subu\ra\Tyu \\
  & \blank[\blank] && : \Tmu\ra\Subu\ra\Tmu \\
  & \blank \circ \blank && : \Subu\ra\Subu\ra \Subu
  \end{alignat*}
These can be defined either by iterating unary substitutions, or by recursion
on untyped syntax: the two ways yield provably equal definitions. In the
following, we assume that it is defined by recursion. We also make use of the
following definition:
  \begin{alignat*}{5}
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \keep && : \Subu \ra \Subu \\
   &&& := \lambda \sigma. (\varu\, 0 \,\consu\, \wkO\,\sigma)
  \end{alignat*}
  The idea is that if $\sigma$ is a substitution between contexts $\Gamma$ and
  $\Delta$, then $\keep\,\sigma$ is a substitution between contexts $\Gamma\ext A[\sigma]$
  and $\Delta\ext A$ for any type $A$. This occurs when defining
  $(\Piu\,A\,B)[\sigma]$ as $\Piu\,(A[\sigma])\,(B[\keep \,\sigma])$.

  We can define the identity substitution on a context $\Gamma$ as
  follows, where $\keep^{\length{\Gamma}}$ is $\keep$ iterated
  $\length{\Gamma}$ times:
\begin{alignat*}{5}
  & \idu && : \Conu \ra \Subu \\
  & && := \lambda \Gamma.\keep^{\length{\Gamma}} \epsilonu
  \end{alignat*}

@@

  \subsubsection{Typing judgments}
  \label{sec:typing_judgments}
  The typing judgments are defined as the following inductive datatype indexed over the
  untyped syntax:
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \blank \vdash && : \Conu\ra \Set \\
  & \blank \vdash \blank  && : \Conu\ra \Tyu\ra\Set \\
  & \blank \vdash \blank \in \blank  && : \Conu\ra \Tmu\ra \Tyu\ra \Set \\
  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
  & \blank \vdash \blank \Rightarrow \blank  && : \Conu\ra \Subu \ra \Conu\ra \Set \\
  % & \Subpre  && : \Set \\
  & \emptyw && : \emptyu \vdash \\
  & \epsilonw && : \Gamma\vdash \epsilonu \Rightarrow \emptyu \\
  & \blank\extw\blank && : ( \Gamma\vdash )  \ra  (\Gamma\vdash A)  \ra
  \Gamma\extu A \vdash \\
  & \consw && :
    (\Delta \vdash) \ra
    (\Gamma\vdash \sigma \Rightarrow\Delta)\ra
    ( \Delta \vdash A) \ra
    ( \Gamma \vdash t \in A[\sigma]) \ra
    \Gamma \vdash t \consu \sigma \Rightarrow \Delta\extu A
   \\
  & \varw  && : (\Gamma \vdash n \invar A) \ra \Gamma \vdash \varu n \in A \\
  & \vzw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra\Gamma \extu  A \vdash 0 \invar \wku \, A \\
  & \vsw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra(\Gamma \vdash n \invar A) \ra (\Gamma \vdash B) \ra \Gamma \extu  B
  \vdash S\,n \invar \wku \, A \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uw && : (\Gamma \vdash)\ra \Gamma \vdash \Uu \\
  & \Elw && : (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra\Gamma \vdash \Elu\,a \\
  & \rlap{(3) Inductive parameters} \\[0.5em]
  & \Piw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
  \Elu\,a \vdash B)\ra\Gamma \vdash \Piu \, a \, B \\
  & \appw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
    \Elu\,a \vdash B)
    \\
    & &&
    \ra(\Gamma \vdash t \in \Piu \, a \, B )
    \ra
    (\Gamma \vdash u \in \Elu\,a)
    \ra
    \Gamma \vdash t \appu u \in  B [0 := u] \\
  & \rlap{(4) Inductive parameters} \\[0.5em]
  & \Pimw && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t) \ra
    \Gamma \vdash \Pimu \,T\,A
    \\
    & \appmw  && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t)
    \\
    & &&
    \ra(\Gamma \vdash t \in \Pimu \,T\,A)
    \ra (u : T) \ra \Gamma \vdash t\,\appmu\,u \in A\,u
\end{alignat*}
There is possibility of redundancy in the arguments of the constructors. Here,
we are ``{paranoid}'', so that we get more inductive hypotheses when performing
recursion.
